---
title: "Beginings"
author: "Al"
date: "2026-01-06"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R programming Course notes

Following the first course beginings. They are telling me how to begin to understand the syntax of R. For example

```{r}
x <- 1 ## <- assignment of value to a variable. This line is assigning the value 1 to the variable x
print(x) ## outputs in the terminal the value of x
x ## Auto-prints the value of x

msg <- "Hello" ## Assigning a string
msg

x <- 1:20 ## Creating a vector populated with the values from 1 to 20 in their respective positions.
x
```

## Data Types

R has 5 basic or "atomic" classes of objects:

- character  
- numeric  
- integer  
- complex  
- logical (boolean)  

## Objects 
The most basic object is a vector

- A vector can only contain objects of the same class  
- *list* is a vector that can contain different classes  

Empty vector is created by the vector() function.

## Numbers

Numbers in R are always treated as doubles. If you want an integer specifically you need to use the L suffix (ex. 1 is double, 1L is an integer). Inf is a special number which represents infinity. Also represented as (1/0). NaN represents an undefined value or missing number.

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
