---
title: "Beginings"
author: "Al"
date: "2026-01-06"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R programming Course notes

Following the first course beginings. They are telling me how to begin to understand the syntax of R. For example

```{r}
x <- 1 ## <- assignment of value to a variable. This line is assigning the value 1 to the variable x
print(x) ## outputs in the terminal the value of x
x ## Auto-prints the value of x

msg <- "Hello" ## Assigning a string
msg

x <- 1:20 ## Creating a vector populated with the values from 1 to 20 in their respective positions.
x
```

## Data Types

R has 5 basic or "atomic" classes of objects:

- character  
- numeric  
- integer  
- complex  
- logical (boolean)  

## Objects 
The most basic object is a vector

- A vector can only contain objects of the same class  
- *list* is a vector that can contain different classes  

Empty vector is created by the vector() function.

## Numbers

Numbers in R are always treated as doubles. If you want an integer specifically you need to use the L suffix (ex. 1 is double, 1L is an integer). Inf is a special number which represents infinity. Also represented as (1/0). NaN represents an undefined value or missing number.

## Attributes

R objeccts can have attributes
- names,dimnames  
- dimensions  
- class  
- length  
- other user-defined attributes/metadata  

Attributes of an objeect can be accessed using the attributes() functions.

## Creating Vectors

The c() function can be used to create vectors of objects

```{r}
x <- c(.5,.6) ## numeric
x
x<- c(TRUE, TRUE) ## boolean
x 
x <- c(T,F)
x
x <- c("a","b","c")
x
x <- 9:29 ## integer
x
x <- c(1+0i,2+4i) ## Complex
x
```

Or using the vector() function

```{r}
x <- vector("numeric", length = 10)
x
```

What about mixing objects?

```{r}
y <- c(1.7,"a") ## char
y
y <- c(T,2) ##numeric
y
y<- c("a", TRUE) ##Char
y
```
They get coerced to the common denominator class

## Explicit coercion

Whay if you want to coerced something instead of letting the machine assume it? You can use the as."" () function.

```{r}
x <- 0:6
class(x)
as.numeric(x)
class(as.numeric(x))
as.logical(x)
as.character(x)
```

Nonsensical coercion results in NA s

## List

Vectors with different classes of opbjects

```{r}
x <- list(1,"a",TRUE,1+4i,2+0i)
x
```


